Supplementary material

### 2022-02-07 analysis/paper/paper.Rmd
Using IDW for the interpolation of the displacement curves. Re-running analyses

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UofO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

@@

```{r get-data-functions, include = FALSE}
# Load all sites and radiocarbon data
sites <- read.csv(here('analysis/data/raw_data/sites.csv'))
rcarbon <- read.csv(here('analysis/data/raw_data/radiocarbon.csv'))

@@

The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shorelines, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to the two phenomena.

To take account of the gradient in the isostatic rebound, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done for each site by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the distance between the isobases. To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. These values were then subtracted from the displacement curve located to the south-west of the site, and in the few cases where the site was located south-west of all isobases, the values were instead added. The result of this process is shown for an example site in Figure \@ref(fig:example-1).

For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase. In the case where there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

@@

```{r example-site, fig.show='hide'}

# Assign closest isobases to site data.
sites_sa <- st_join(st_make_valid(sites_sa), isopolys,
                    join = st_intersects, largest = TRUE)

# Specify site name
sitename <- "Hegna vest 1"

@@

# Find local displacement curve
sitecurve <- interpolate_curve(years = xvals,
                                 isobase1 = sitel$isobase1,
                                 isobase2 = sitel$isobase2,
                                 target = sitel,
                                 dispdat = displacement_curves,
                                 isodat = isobases,
                                 direction_rel_curve1 = sitel$dir_rel_1)
# Add site name
sitecurve$name <- sitename

@@

dispdata <- filter(displacement_curves, name %in% c("Larvik", "Tvedestrand"))

curve_plot <- ggplot() + 
              geom_line(data = dispdata, 
                        aes(x = years, y = upperelev, col = name)) +
  geom_line(data = dispdata, 
                        aes(x = years, y = lowerelev, col = name)) +
  geom_line(data = sitecurve,
            aes(x = years, y = upperelev, col = name)) +
  geom_line(data = sitecurve,
            aes(x = years, y = lowerelev, col = name)) +
  scale_colour_manual(values = c("Larvik" = "darkgreen",
                                  "Tvedestrand" = "blue",
                                 "Hegna vest 1" = "red")) +
  ylab("Meters above present sea-level") +
  xlab("cal BCE/CE") +
  theme_bw() +
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal")

@@

```{r example-1, echo = FALSE, fig.cap = "Example site Hegna vest 1 (Fossum 2017). A) Location of the site in the present day landscape. The red outline is the site limit. B) Radiocarbon dates associated with the site. Fill colour indicates what dates are assumed to belong to the same settlement phase. Multiple dates are modelled using the Boundary function in OxCal and then summed. The red outline indicates that the date does not match the typological indicators in the artefact assemblage of the site. C) The location of the site within the study area relative to isobases of the employed displacement curves. D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/example_map.png"))
```

```{r example-site-cont2, fig.show='hide'}
set.seed(1)

# Retrieve example date to be used. The posterior sum of date group 1
posteriorprobs <- filter(datedat, group == 1 & class == "sum")

@@

```{r redate, echo = FALSE, fig.cap = "95% probability ranges of shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories. These are comapred to the dates originally proposed by the excavation reports in black.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/redate.png"))
```

>>>>

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

@@

```{r get-data-functions, include = FALSE}
# For reproducibility
set.seed(123)

# Load all sites and radiocarbon data
sites <- read.csv(here('analysis/data/raw_data/sites.csv'))
rcarbon <- read.csv(here('analysis/data/raw_data/radiocarbon.csv'))

@@

The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shorelines, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to the two phenomena. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was interpolated to each site location from the displacement curves using inverse distance weighting based on the distance between the site and the isobases of the displacement curves [e.g. @conolly2020; @conolly2006,94--97]. This was done for each year along the entirety of the curves using the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase. In the case where there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

@@

```{r example-site, fig.show='hide'}
# Specify site name
sitename <- "Hegna vest 1"

@@

# Find local displacement curve
sitecurve <- interpolate_curve(years = xvals,
                                 target = sitel,
                                 dispdat = displacement_curves,
                                 isodat = isobases)
# Add site name
sitecurve$name <- sitename

@@

curve_plot <- ggplot() + 
              geom_line(data = displacement_curves, 
                        aes(x = years, y = upperelev, col = name)) +
  geom_line(data =  displacement_curves, 
                        aes(x = years, y = lowerelev, col = name)) +
  geom_line(data = sitecurve,
            aes(x = years, y = upperelev, col = name)) +
  geom_line(data = sitecurve,
            aes(x = years, y = lowerelev, col = name)) +
  scale_colour_manual(values = c("Arendal" = "black",
                                   "Larvik" = "darkgreen",
                                   "Tvedestrand" = "blue",
                                   "Horten" = "darkorange",
                                 "Hegna vest 1" = "red")) +
  ylab("Meters above present sea-level") +
  xlab("cal BCE/CE") +
  theme_bw() +
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal")

@@

```{r example-1, echo = FALSE, fig.cap = "Example site Hegna vest 1 (Fossum 2017). A) Location of the site in the present day landscape. The red outline is the site limit. B) Radiocarbon dates associated with the site. Fill colour indicates what dates are assumed to belong to the same settlement phase. Multiple dates are modelled using the Boundary function in OxCal and then summed. The red outline indicates that the date does not match the typological indicators in the artefact assemblage of the site. C) The location of the site within the study area relative to isobases of the employed displacement curves. D) Displacement curve interpolated to the site location.", out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/example_map.png"))
```

```{r example-site-cont2, fig.show='hide'}
# Retrieve example date to be used. The posterior sum of date group 1
posteriorprobs <- filter(datedat, group == 1 & class == "sum")

@@

```{r redate, echo = FALSE, fig.cap = "95\\% probability ranges of shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories. These are comapred to the dates originally proposed by the excavation reports in black.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/redate.png"))
```

### 2022-02-07 analysis/supplementary-materials/supplementary.Rmd
Using IDW for the interpolation of the displacement curves. Re-running analyses

>>>>

```{r auve,  echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/anvik.png"))

filter(rcarbon, site_name  == "Auve") %>% 
  select(lab_code, c14_bp, error, material, context) %>% 
   kable(caption = "Auve", col.names = c("ID",
                    "\\textsuperscript{14}C BP",
                    "Error",
                    "Material",
                    "Context"),
        booktabs = TRUE,  linesep = c(""), escape = FALSE)

The site limit for Auve was drawn manually based on the description in the publication of the site and the site geometry available in the database Askeladden. The area to the has reportedly been affected by land-slides, but the extent of this is difficult to ascertain based on maps of the area. As the simulation results appear reasonable, nothing was done to try to correct for this.
